---
title: "PA1_template"
author: "treepruner"
date: "November 2, 2015"
output: html_document
---

rm(list = ls())

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, echo = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(sqldf)
library(lubridate)
library(mosaic)
```

#### Overview

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.


The variables included in this dataset are:

 * steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

 * date: The date on which the measurement was taken in YYYY-MM-DD format

 * interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.


#### Get Data

```{r get_data, echo = TRUE, message = FALSE}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileURL, "./proj1/repdata%2Fdata%2Factivity.zip", method = "curl")
unzip(zipfile = "repdata%2Fdata%2Factivity.zip")
dir()
activity <- read.csv("./proj1/activity.csv")

```

#### Exploratory Analyses

Display the size of the data frame and some sample records that aren't NA.

 * look at summaries
 * check for missing data
 * create exploratory plots
 * perform exploratory analyses 

```{r explor_data, echo = TRUE}
activity_rowCnt <-nrow(activity)
str(activity)
head(activity, n = 5)
tail(activity, n = 5)
activity[5000:5005,]

summary(activity)
```


```{r date_steps_plot, echo = TRUE}
plot(activity$date, activity$steps)
```



#### Clean and Preprocess Data

The date is currently a factor and needs to be converted to a date.Use dplyr and mosaic packages to create a new factor variable to indicate weekend or weekday.


```{r date_cleanup, echo = TRUE}
activity$modDate <- as.Date(activity$date)

weekend <-c("Saturday", "Sunday")

activity <- mutate(activity,
  dayType = derivedFactor(
 "weekend" = (weekdays(activity$modDate) %in% weekend),
 .method = "first",
 .default = "weekday"
))
```


Retain only the completed cases in a new data frame to resolve the NA issue. 

```{r incomplete_cases, echo = TRUE}
cc <-activity[complete.cases(activity),]
cc_rowCnt <- nrow(cc)
incomplete <- activity_rowCnt - cc_rowCnt
```


#### What is mean total number of steps taken per day for complete cases?

dplyr is the most understandable and easiest way to do this:

```{r cc_stepsByDate, echo = TRUE }
stepsByDate <- 
        cc %>% 
        group_by(modDate) %>% 
        summarise(n_date = n(), sum_dateSteps = sum(steps) , mean_dateSteps = mean(steps))
```


#### Histogram of Complete Case Steps by Date

```{r cc_stepsByDate_histogram, echo=TRUE, fig.height = 3.5, fig.width = 3.5}
hist(stepsByDate$sum_dateSteps,
     xlab = "Steps",
     breaks = 10,
     main = "Histogram of Complete Cases Daily Steps"
     )
```


```{r cc_stats, echo = TRUE }
datasetMean <- mean(stepsByDate$sum_dateSteps)
datasetMedian <- median(stepsByDate$sum_dateSteps)
```

The mean of the total number of steps for complete cases taken per day is `r as.character(signif(datasetMean,7))`.

The median of the total number of steps for complete cases taken per day is `r as.character(signif(datasetMedian,7))`.

### What is the average daily activity pattern for complete cases?

Use dplyr to create summaries by 5 minute interval:


```{r cc_stepsByInterval, echo = TRUE }
stepsByInterval <- 
        cc %>% 
        group_by(interval) %>% 
        summarise(n_interval = n(), sum_intervalSteps = sum(steps) , mean_intervalSteps = mean(steps))
```



#### Times Series Plot

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r cc_interval_stats, echo = TRUE}
stepsByInterval[which(stepsByInterval$mean_intervalSteps == max(stepsByInterval$mean_intervalSteps)),]

maxInterval <- stepsByInterval[which(stepsByInterval$mean_intervalSteps == max(stepsByInterval$mean_intervalSteps)),1]

```


```{r cc_interval_plot, echo = TRUE, fig.height=3.5, fig.width = 3.5 }
plot(stepsByInterval$interval, stepsByInterval$mean_intervalSteps,
    type = "l",
    ylab = "Mean Steps",
    xlab = "Interval")
    abline(v = maxInterval, col = "red")
 

```


### Impute Missing Values

The original dataset had `r activity_rowCnt` rows. 
The data set after removing the incomplete cases had `r cc_rowCnt` rows.

This is a difference of 'r incomplete` records

The NA records in the original data set have been defaulted to use the mean for the particular 5 minute interval and saved as the data frame activityImputed. The package sqldf was used to join to the interval summary dataframe.



```{r impute, echo = TRUE }
activityImputed1 <- sqldf( 
        "select a.*, i.mean_intervalSteps, i.sum_intervalSteps    
        from activity a join stepsByInterval i on a.interval = i.interval")

activityImputed2 <- sqldf(c("update activityImputed1 set steps = mean_intervalSteps where steps  is null", "select * from main.activityImputed1"))

summary(activityImputed2)

```


#### Summarize the Imputed File

```{r imputedStepsByDate, echo = TRUE }
imputedStepsByDate <- 
        activityImputed2 %>% 
        group_by(modDate) %>% 
        summarise(n_date = n(), sum_dateSteps = sum(steps) , mean_dateSteps = mean(steps))

```


Make a histogram of the steps by day including the imputed values

```{r imputedStepsByDate_histogram, echo=TRUE, fig.height = 3.5, fig.width = 3.5}
hist(imputedStepsByDate$sum_dateSteps,
     xlab = "Steps",
     breaks = 10,
     main = "Histogram of Daily Steps with Imputed Values"
     )
```



```{r imputedStepsByDate_stats, echo = TRUE }
imputedDatasetMean <- mean(imputedStepsByDate$sum_dateSteps)
imputedDatasetMedian <- median(imputedStepsByDate$sum_dateSteps)

mean_diff <- datasetMean - imputedDatasetMean
median_diff <- datasetMedian - imputedDatasetMedian
```

The mean of the total number of steps   taken per day is `r as.character(signif(imputedDatasetMean,7))` after imputation.

The median of the total number of steps  taken per day is `r as.character(signif(imputedDatasetMedian,7))` after imputation.

The mean changed by `r as.character(signif(mean_diff,7))`

The median changed by `r as.character(signif(median_diff,7))`


### Are there differences in activity patterns between weekdays and weekends?


Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.


```{r imputedStepsByDayType, echo = TRUE }
imputedStepsByDayType <- 
        activityImputed2 %>% 
        group_by(dayType) %>% 
        summarise(n_date = n(), sum_dayTypeSteps = sum(steps) , mean_dayTypeSteps = mean(steps))

```




#### Panel Plot of Avg Steps Taken 


```{r imputed_dayType_plot, echo = TRUE }
xyplot(imputedStepsByDayType$interval ~  imputedStepsByDayType$mean_intervalSteps | imputedStepsByDayType$dayType,
    type = "l",
    ylab = "Mean Steps",
    xlab = "Interval")


```

