---
title: "Health and Economic Impacts of US Storms"
author: "treepruner"
date: "November 15, 2015"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

rm(list = ls())

```{r load_packages, echo = TRUE, message = FALSE}
library(lubridate)
library(dplyr)
library(sqldf)
library(xtable)
library(ggplot2)


library(mosaic)
```



### Synopsis

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

Additional inforation can be found at these websites:

https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf


 [1] "STATE__"    "BGN_DATE"   "BGN_TIME"   "TIME_ZONE" 
 [5] "COUNTY"     "COUNTYNAME" "STATE"      "EVTYPE"    
 [9] "BGN_RANGE"  "BGN_AZI"    "BGN_LOCATI" "END_DATE"  
[13] "END_TIME"   "COUNTY_END" "COUNTYENDN" "END_RANGE" 
[17] "END_AZI"    "END_LOCATI" "LENGTH"     "WIDTH"     
[21] "F"          "MAG"        "FATALITIES" "INJURIES"  
[25] "PROPDMG"    "PROPDMGEXP" "CROPDMG"    "CROPDMGEXP"
[29] "WFO"        "STATEOFFIC" "ZONENAMES"  "LATITUDE"  
[33] "LONGITUDE"  "LATITUDE_E" "LONGITUDE_" "REMARKS"   
[37] "REFNUM"   



### Executive summary
This analysis will address these questions: 

 * Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

 * Across the United States, which types of events have the greatest economic consequences?

### Data Processing

#### Load Data



```{r getData, echo = TRUE}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

download.file(fileURL, "./proj2/repdata%2Fdata%2FStormData.csv.bz2")

stormData <- read.csv("./proj2/repdata%2Fdata%2FStormData.csv.bz2")
dir()

# save original
saveRDS(stormData, "./proj2/stormData.rds")

str(stormData)
names(stormData)
head(stormData,1)
envTypeList <- unique(stormData$EVTYPE) ## 985 types
write.table(envTypeList, file = "./proj2/envTypeList.csv", col.names = TRUE)
```


```{r reGetData, echo = TRUE}
sd <- readRDS("./proj2/stormData.rds")

```



```{r saveSampleData, echo = TRUE}

# save sample to work with
sdSample <- stormData[sample(nrow(stormData), 2000), ]

# save sample
saveRDS(sdSample, "./proj2/sdSample.rds")

str(sdSample)
names(sdSample)
head(sdSample,1)
envTypeList <- unique(sdSample$EVTYPE) ## 985 types

sd <- sdSample

rm(stormData)

```

```{r GetSampleData, echo = TRUE}
sd <- readRDS("./proj2/sdSample.rds")
```





#### Clean/Preprocess Data
There are 985 unique EVTYPE codes, which is too large to use. The EVTYPE codes were groupd together into 38 groups. The new EnvTypeGroup variable was added to the dataset with the sqldf package.

```{r addEnvTypeGroup, echo = TRUE}
envTypeList <- read.csv("./proj2/envTypeList.csv",  sep = ",")


sd <- sqldf( 
        "select s.*, e.envTypeGroup   
        from sd s join envTypeList e on s.EVTYPE = e.ENVTYPE")

```

Use lubridate to convert the dates that are factors into date fields.

```{r date_cleanup, echo = TRUE}

sd$BGN_DATE <- mdy_hms(sd$BGN_DATE)
sd$END_DATE <- mdy_hms(sd$END_DATE)

```


Determine if the REFNUM identifies a unique row. Remove duplicate REFNUMs. This removeds records were exactly the same in all fieds. There is another set of records, REFNUM 605943 and 567221 that although they have different REFNUMs, appear to be for the same event. The REMARKS refer to millions in damage, so REFNUM 605943 which referred to billions, was removed.

```{r dup_refnum_check, echo = TRUE}
dup_refnum <- sqldf( 
        "select s.REFNUM, count(*)   
         from sd s 
         group by s.REFNUM
         having count(*) > 1")

dup_refnum

sd <- sqldf( 
        "select distinct s.*   
         from sd s" 
         )

# thes code also works and was actually faster
# sd1 <- sd[!duplicated(sd),] 



```

The PROPDMG field is the numerical component of the property damange and the PROPDMGEXP is the units, however, the units field is dirty. The numerical field was multiplied as follows: 
 * Bb = billions
 * Mm = millions 
 * Kk = thousands
 * Hh = hundreds
 * ?  = 0
 * anything else = millions
 
The CROPDMG field is the numerical component of the crop damange and the CROPDMGEXP is the units, however, the units field is dirty. The numerical field was multiplied as follows: 
 * Bb = billions
 * Mm = millions 
 * Kk = thousands
 * Hh = hundreds
 * ?  = 0
 * anything else = millions

The events in this file range from 1950 to 2011. To truely be comparable and understandable to the modern audience, the damage dollar amounts should be adjusted using an appropriate price index into current dollars in order to properly rank the financial impact of each event.

```{r fix_amts, echo = TRUE}
sd1 <- sqldf( 
        "select  REFNUM
        , BGN_DATE
        , END_DATE
        , EVTYPE
        , EnvTypeGroup
        , FATALITIES
        , INJURIES
        , PROPDMG
        , PROPDMGEXP
        , case 
                when upper(PROPDMGEXP) = 'B' then PROPDMG * 1000000000
                when upper(PROPDMGEXP) = 'M' then PROPDMG * 1000000
                when upper(PROPDMGEXP) = 'K' then PROPDMG * 1000
                when upper(PROPDMGEXP) = 'H' then PROPDMG * 100
                when upper(PROPDMGEXP) = '?' then PROPDMG * 0
           else PROPDMG * 1000000
           end as PropertyDamageAmt
        , CROPDMG
        , CROPDMGEXP
        , case 
                when upper(CROPDMGEXP) = 'B' then CROPDMG * 1000000000
                when upper(CROPDMGEXP) = 'M' then CROPDMG * 1000000
                when upper(CROPDMGEXP) = 'K' then CROPDMG * 1000
                when upper(CROPDMGEXP) = 'H' then CROPDMG * 100
                when upper(CROPDMGEXP) = '?' then CROPDMG * 0
           else CROPDMG * 1000000
           end as CropDamageAmt
        , REMARKS   
         from sd
         where REFNUM <> 605943 "          )

saveRDS(sd, "./proj2/sd.rds")
rm(sd)
```         
         


```{r pop_stats, echo = TRUE}
pop_fatalities <- 
        sd1 %>% 
        group_by(EnvTypeGroup) %>% 
        summarise(n_date = n()
                  , Fatalities.Count = sum(FATALITIES) 
                  , Fatalities.Mean = mean(FATALITIES)
                  , Fatalities.Median = median(FATALITIES)
                  , Fatalities.Min = min(FATALITIES)
                  , Fatalities.Max = max(FATALITIES)        ) %>% 
        arrange(desc(Fatalities.Count))



pop_injuries <- 
        sd1 %>% 
        group_by(EnvTypeGroup) %>% 
        summarise(n_date = n()
                  , Injuries.Count = sum(INJURIES) 
                  , Injuries.Mean = mean(INJURIES)
                  , Injuries.Median = median(INJURIES)
                  , Injuries.Min = min(INJURIES)
                  , Injuries.Max = max(INJURIES)    ) %>% 
        arrange(desc(Injuries.Count))
        



```


```{r HealthImpactTable, echo = TRUE,  results="asis"}
fatalities <- xtable(pop_fatalities )
print(fatalities, type="html")


injuries <- xtable(pop_injuries )
print(injuries, type="html")

```


```{r pop_stats, echo = TRUE}
cost_PropertyDamage <- 
        sd1 %>% 
        group_by(EnvTypeGroup) %>% 
        summarise(n_date = n()
                  , PropertyDamageAmt.Sum = sum(PropertyDamageAmt) 
                  , PropertyDamageAmt.Mean = mean(PropertyDamageAmt)
                  , PropertyDamageAmt.Median = median(PropertyDamageAmt)
                  , PropertyDamageAmt.Min_ = min(PropertyDamageAmt)
                  , PropertyDamageAmt.Max = max(PropertyDamageAmt)   ) %>% 
        arrange(desc(PropertyDamageAmt.Sum))


cost_CropDamage <- 
        sd1 %>% 
        group_by(EnvTypeGroup) %>% 
        summarise(n_date = n()
                  , CropDamageAmt.Sum = sum(CropDamageAmt) 
                  , CropDamageAmt.Mean = mean(CropDamageAmt)
                  , CropDamageAmt.Median = median(CropDamageAmt)
                  , CropDamageAmt.Min = min(CropDamageAmt)
                  , CropDamageAmt.Max = max(CropDamageAmt)   ) %>% 
        arrange(desc(CropDamageAmt.Sum))


```


```{r cost_table, echo = TRUE,  results="asis"}
PropertyDamage <- xtable(cost_PropertyDamage )
print(PropertyDamage, type="html")


CropDamage <- xtable(cost_CropDamage )
print(CropDamage, type="html")
```







#### 

```{r, echo = TRUE }

```




#### 


```{r , echo=TRUE}

```




```{r, echo = TRUE }

```



#### 




#### 

```{r, echo = TRUE }

```





#### 


```{r, echo = TRUE }

```



```{r, echo = TRUE }

```



#### 



```{r, echo = TRUE }

```



#### 


```{r, echo = TRUE }

```

### Results

 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

 2. Across the United States, which types of events have the greatest economic consequences?